---
title: "Microbiome BMI analyses - regression models"
subtitle: "Main analysis and three sensitivity analyses"
author: "Marlena Maziarz & Mariam Miari"
date: "`r format(Sys.Date(), '%B %e, %Y')`"
output:
  bookdown::html_document2:
    toc: yes
    toc_depth: 5
    toc_float: yes
    fig_caption: yes
  bookdown::pdf_document2:
    toc: yes
    toc_depth: 5
    fig_caption: yes
  bookdown::word_document2:
    toc: yes
    toc_depth: 5
    fig_caption: yes
always_allow_html: true
editor_options: 
  chunk_output_type: inline
---

# The main analysis

## Setup
Load the data, source the helper functions etc. The dataset was prepared by Mariam, filtered at 5% and clr-transformed, n = 9993 where I used unrelated individuals (UNRELATED_1_DEGREE) 

```{r}
rm(list = ls())
library(tidyverse)
library(rio)
library(here)
library(robustbase)
library(readxl)

results_version <- '240906'
##################################################################################################################################
# Source the helper functions
source(here::here('bmi_paper_model_fitting_helper_functions_240906.R')) # NOTE THE VERSION, make sure it's the right one!
##################################################################################################################################



# reads 'data' 
load("/Volumes/research/metabogut/BMI_2021/data/processed/MGS2.0/data_nds_filt_CLRtrans_9993_updated250208.RData")
data <- data_nds_trans_filt

# Since the regression model does not accept spaces in the variables names being regressed I will take the mgs ID only and run the regression model with it

matching_columns <- grepl("hMGS", colnames(data))

# Replace everything before "hMGS" with an empty string
colnames(data)[matching_columns] <- sub(".*hMGS\\.", "hMGS.", colnames(data)[matching_columns])


data <- data
#factor physical_activity
data <- data %>%
  mutate(physical_activity_f = factor(physical_activity)) %>%
  relocate(physical_activity_f, .after = physical_activity)

#factor smoking
data <- data %>%
  mutate(smoking_f = factor(smoking)) %>%
  relocate(smoking_f, .after = smoking)

data <- data %>% 
  mutate(male = sex == 1)
```



```{r}
model_names <- c('M1_BMI',
                 'stratM1_BMI_females',
                 'stratM1_BMI_males',
                 'intM1_BMI')

# covariates for each of the base models
base_model_covariates_in_all_models <- c('age', 'country_birth', 'Cohort')

base_model_covariates <- list(
  M1_BMI                       = c(base_model_covariates_in_all_models, 'male'),
  stratM1_BMI_females          = c(base_model_covariates_in_all_models),
  stratM1_BMI_males            = c(base_model_covariates_in_all_models),
  intM1_BMI                    = c(base_model_covariates_in_all_models, 'male'))

# covariates for each of the fully adjusted models
full_model_covariates_in_all_models <- c('age', 'country_birth', 'Cohort', 'physical_activity_f', 'smoking_f', 'metformin', 'diab_diag', 'Fiber')

full_model_covariates <- list(
  M1_BMI                       = c(full_model_covariates_in_all_models, 'male'),
  stratM1_BMI_females          = c(full_model_covariates_in_all_models),
  stratM1_BMI_males            = c(full_model_covariates_in_all_models),
  intM1_BMI                    = c(full_model_covariates_in_all_models, 'male'))

model_outcome <- list(
  M1_BMI                       = 'BMI',
  stratM1_BMI_females          = 'BMI',
  stratM1_BMI_males            = 'BMI',
  intM1_BMI                    = 'BMI')

model_with_interaction <- list(
  M1_BMI                       = F,
  stratM1_BMI_females          = F,
  stratM1_BMI_males            = F,
  intM1_BMI                    = T)

model_strata <- list(
  M1_BMI                       = 'all',
  stratM1_BMI_females          = 'females',
  stratM1_BMI_males            = 'males',
  intM1_BMI                    = 'all')
```


## Model fitting

In this part we use all the data, ie. data from the three cohorts and adjust for cohort.

```{r}
all_results_main <- NULL

for(curr_model in model_names){
    
    curr_results <- get_results_for_bmi_paper(d = data,
                                              outcome = model_outcome[[curr_model]], 
                                              model_name = curr_model,
                                              model_strata = model_strata[[curr_model]],
                                              suppress_printing = F,
                                              base_model_vars_to_adjust_for = base_model_covariates[[curr_model]],
                                              full_model_vars_to_adjust_for = full_model_covariates[[curr_model]],
                                              glm_family = 'gaussian', # with glmrob() method = "Mqle" is added
                                              mgs_sex_interaction = model_with_interaction[[curr_model]],
                                              sex_var = 'male')
    
    export(curr_results, file = here::here('results', 'main_analyses', paste0(curr_model, '_results_', results_version, '.xlsx')))
    all_results_main <- bind_rows(all_results_main, curr_results)
    
    rm(curr_results)
}

all_results_main <- all_results_main %>%
    mutate(analysis = 'main',
           cohort = 'all_cohorts') 

```

```{r}
all_results_main$analysis <- ifelse(grepl("base_M1_BMI|full_M1_BMI", all_results_main$model_name), "all", 
                                    ifelse(grepl("base_stratM1_BMI_females|full_stratM1_BMI_females|base_stratM1_BMI_males|full_stratM1_BMI_males",
                                                 all_results_main$model_name), "sex_stratification",
                                           "interaction"))
```

# FDR correction
## for interaction, do the FDR adjustment per interaction effect
```{r}
interaction_df <- all_results_main %>%
    filter(grepl("int", model_name) & grepl(":maleTRUE", mgs))

interaction_df$p_value_adj <- p.adjust(interaction_df$p, method = "fdr") # correct interaction results alone (int effect only)

interaction_df$outcome <- "BMI"
interaction_df <- interaction_df[,c("analysis", "outcome","cohort","model_name","mgs","n_obs","est","se","ci_low","ci_high",
                      "p", "p_value_adj")]

rest_df <- all_results_main %>%
  filter(!grepl("int", model_name))

results <- rest_df %>%
  group_by(analysis) %>%
  mutate(p_value_adj = p.adjust(p, method = "fdr")) %>%
  ungroup()

results$outcome <- "BMI"
results <- results[,c("analysis", "outcome","cohort","model_name","mgs","n_obs","est","se","ci_low","ci_high",
                      "p", "p_value_adj")]

all_results <- rbind(interaction_df, results)

```

## save
```{r}
export(all_results, here::here("results/results_240906.xlsx"))
```

